draft 1

### 2016-07-04 finalwriteup/draft.rmd
gave draft to kerrrie

opts_chunk$set(fig.width=8, fig.depth = 8, dpi = 300, warning = FALSE, message = FALSE, echo = FALSE)
```

@@

# Contents

# List of Figures

@@

**ANOVA** - Anaylis of Variance. A statistical test that can be applied to find whether there is a significant difference between the means of more than two groups.

@@

**ROC** - Reciever Operating Characteristic. An ROC curve plots the performance of a binary classification algorithm its scoring threshold is varied.

@@

@Lovallo2003 call the practise of analysing a project based on project details and its unique complexities the 'inside view'. This constrasts with an 'outside view' where a subject is analysed by comparison to other subjects and disregarding specific subject details. For example, research has shown that if people make predictions about their skills after being exposed to an 'outside view' (a summary of other people's skills), their predictions are significantly more accurate. According to @Lovallo2003, this 'outside view' can be applied to complex projects, and stipulates that the details of the current project are ignored in favour of analysing outcomes of several similar projects. Mobilising the outside view in this way is called reference class forecasting [@Lovallo2003]. One method @Lovallo2003 recommend is to obtain correlation statistics from past similar projects - the correlation between the forecast cost and the actual cost. The correlation for the current project can then be estimated via a statistical model, which is then used to adjust the forecasted cost made by detailed analysis (the inside view) [@Flyvbjerg2011]. Managers rarely analyse data from similar projects even though it could significantly improve cost predictions [@Flyvbjerg2011]. This tendency led to a formalised introduction of 'outside view' methods into some institutions.

@@

In contrast to construction, IT projects are a relatively new practise where alternative contractural arangements have been trialled. Through experimentation with contracts, the Agile movement was developed. In summary, the method treats both cost and time as fixed quantities for a project, and any change or additional work can only be accommodated if another, less important requirement is excluded [@Badenfelt2011]. Furthermore, the project outcomes are continually adjusted ad revised based on frequent progress review points that assesses unexpected problems. This agile attitude has become ingrained in many IT practices and the flexible upfront arrangement with clients often forges better long term relationships [Cockburn2001].

@@

A statistical or machine learning model predicting a measure of project financial success was developed using the available project information and by trialling numerous machine learning algorithms and statistical methods. The company uses well established CRM software which readily provides simple output statistics, however, analytical capabilities were limited to:

@@

Fixed price cost estimation for complex projects is a difficult task and when performed incorrectly, can have devastating affects on businesses and governments. Many theories have delved into human nature and our tendency to optimistically assess our capabilities with respect to a given task. In this case, the optimism often results in negative business consequences affecting employee morale and the capacity of a company to flourish and grow. Over the past few decades, significant research has been dedicated to creating predictive models that present the 'outside view' of a project, by statistically comparing a new project to a collection of similar projects and their characteristics. It has been shown that these models improve the cost estimation, however industry uptake has not been successful. It has proven to be a challenging task persuading managers to use reference class forecasting models in practise and remove themselves from summaries based on intricate details of the completed project. This may be due to a number of reasons including lack of understanding, time, and lack of collaboration with industry in development of these models. This research examines the potential for a company's internally generated CRM data to be used in creating a reference class forecasting model. Familiarity with the data has the potential to influence managers' decisions in the company more, however its limitations must be understood in order to create a valid model.

@@

In addition to addressing cost estimation with a different kind of dataset and different industry sector, ensemble tree methods will be applied which have been minimally tested on this problem. To date, Linear Regression, Neural Networks, case based reasoning, and SVM's have been used to predict cost/effort even though ensemble tree methods can perform as well as Neural Networks in some cases and generally outperform Linear Regression. Another benefit of ensemble trees is the output types available, such as partial dependency plots and variable importance plots provide insight into the model's calculations. This is in contrast to Neural Networks and support vector machines (SVM's) which are 'black-box' predictors. Providing insight to a user is a major benefit in this case as past research has demonstrated issues with user trust and uptake of models, even if the models were successful. It is anticipated that using internal data only also improves user trust and creates tangible meaning for users. Blending multiple machine learning and statistical models was trialled which has not yet been applied to the cost estimation problem.

@@

@Elfaki2014 discusses how much detailed cost estimates for construction projects can vary from estimator (engineer) to estimator, which contributes to the lack of accuracy in predciting final cost of a project. He argues that an engineer's expertise is not documented or measured in any way and therefore their expertise and, in turn, their estimate is prone to subjectivity. @Shane2009 theorises that final cost can be subject to so many different unpredictable parameters, such as weather, unexpected ground conditions, duration or sub-contractor issues, that it is almost impossible to achieve an accurate cost prediction manually.

@@

@Elfaki2014 completed a study on cost estimation techniques and research from 2004 to 2014. They found that artificial Neural Networks and SVM's were the most common machine learning techniques used. He highlighted that these two machine learning techniques deal well with uncertainty, however lack technical justification to the decision maker (known as a black box). Neural networks are also time consuming to train, and must be re-trained and re-tested with each additional piece of data [@Kim2004]. Nevertheless, Neural Networks and SVM's recieved a significant amount of attention in the 1990's for their ability to accurately predict construction costs with limited detailed information [@Shin2015][@Kim2004].

@@

Despite Neural Networks typically outperforming Linear Regression models, it is interesting to note some of the outcomes of regression studies in cost estimation. Different studies tended to produce a wide range of variables that contribute at fluctuating levels in the models. This is likely because construction projects are multidisciplinary and involve a range of different parties such as the clients, consultants, contractors and suppliers, each of which contribute varying levels of input [@Chan2005]. @Chan2005's study on 87 building projects in Singapore included special, complex projects. Therefore, variables such as contractor's specialised skills, who the client was (public vs. private), the client level of experience, and the contractor's financial management highly influenced final cost of the project. Other studies found variables such as project complexity, duration, team experience, information availablility, site requirements, and labour climate to contribute most [@Akintoye2000][@Trost2003][@Pinto1988]. As these studies obtained information from questionaires, the available variables also largely depended on the questions in the questionaires, and therefore each study produced a range of different important variables. Despite the differences in qualitative outcomes, the studies all generally reported their mean absolute error or a similar metric from the tested models. Similar result structures are useful, however there is a gap in the literature in how to translate the improved errors from the model into a well justified argument for industry to reduce losses by integrating a parametric model into their business.

@@

1. Storing a collection of projects with key variable values
2. Once a new project or case arrives, similar cases are retrieved. This can be achieved either algorithmically with a distance function applied to the variables, or manually by users reviewing past cases. The nearest neighbour algorithm has been used to find the distance between cases in previous studies. It is programmed to calculate the euclidean distance in n-dimensional space between cases, where each variable is a dimension [@Kumar2007].
3. The cost of the new project is estimated by extrapolating characteristics of the similar cases to the new case. This can also be done either algorithmically or manually by an experienced decision maker.

@@

In summary, research into cost estimation in the construction industry has trialled a variety of methods ranging from detailed manual analysis, algorithmic models, and case based reasoning. Neural networks generally outperformed all other systematic methods, such as regression and case based reasoning, but had the significant disadvantage of being a black box. Alternative methods such as Boosted Trees have the potential to perform as well as Neural Networks and provide insight into the structure of the algorithm. This is true for CBR as well, although it has not been shown to predict as accurately as Neural Networks. There exists a gap in the research in ways to intelligently combine CBR with machine learning methods that could predict well, provide insight to the structure, and engage the user in reviewing similar projects. Also, as stated previously, there exists a gap in the application of cost estimating in the construction industry to smaller components of the project, such as consulting and contracting companies who each face a similar problem in estimating their fixed fee. These companies could mobilise information in their internal project databases, as opposed to the current body of literature that used data from dozens of companies.

@@

Effort estimation is indeed a problem, and was demonstrated by a review of 6 surveys of IT projects between 1984 and 1994 by @Molokken2003. It revealed that 60-80% of IT projects encountered effort and/or schedule overruns, where the average overrun was by 30 to 40%. A survey of software managers asked whether they saw effort estimation as a problem, and 91% responded 'yes', while only 9% answered 'no' [@Moores1992]. The most popular response for the supposed reason for cost overruns was over-optimistic estimates (51%), which falls squarely within the theory of optimism bias, or the 'inside view' [@Flyvbjerg2011]. Given this widespread problem in the industry, plenty of research has been dedicated to the problem of effort prediction, with methods that can be categorised similarly to the construction industry: detailsed analysis, algorithmic modelling, and case based reasoning.

@@

Similar to the construction industry, expert judgment or detailed analysis, is the most widely practised method for effort estimation [@Shepperd1996][@Molokken2003]. This is despite many years of research being dedicated to developing algorithmic models and industry tools that, in the research context, outperform expert judgment. A survey by @Moores1992 found that software tools assisting a software project were in significant use, but only the project planning tools, not the cost estimating tools. This suggests a mental link for project managers to price jobs based on *details* of the software project as opposed to taking an outside view. Another study by @Heemstra1992 found that there was no evidence estimation accuracy improved when estimation tools were used. @Bergeron1992 similarly found jobs that used algorithmic models were actually associated with less accurate estimtes. However, this trend may be coincidental due to a lack of cases where estimation tools were used [@Molokken2003]. The scarcity of real project evidence that algorithmic estimation tools improve estimations and an unintuitive mental jump could be reasons contributing to detailed expert analysis remaining the most widespread technique for effort estimation. This trend mirrors what was found in the construction industry.

@@

As alluded to previosuly, studies on effort estimation as opposed to project cost estimation are concentrated in the software industry. This leaves a gap in studies for effort estimation in other industries such as the construction industry. Infrastructure projects do not have a function point variable and would need to utilise other information such as expected total project cost (which is notoriously hard to predict based on the previous section), client characteristics, and other project characteristics.

@@

In the IT industry, CBR applications have been researched using similar methodologies to the construction industry. The procedure is described as algorithmically finding similar projects using measurements of euclidean distance in n-dimensional space, where each dimension corresponds to a variable [@Shepperd1996]. Then, the similar cases are used to predict effort for the new case. In @Shepperd1996's study, Linear Regression was employed.

@@

Research into effort estimation in the software industry has followed a similar path to cost prediction studies on construction projects. Expert judgment via detailed analysis is still the predominant method for effort/cost estimation. Case based reasoning and algorithmic methods such as Neural Networks have found some success in effort estimation (at similar levels) but have not been successfully adopted by industry. It was noted that case based reasoning resonated more with decision makers because its predcitions could be better justified by the model. The visibility of model structure is very important for future work if the results are to be translated to industry. Effort estimation remains a problem, however agile methods of project management and effort management are changing the contractual approach of consulting in the software industry [@Badenfelt2011]. This innovative method keeps delivery times short (and therefore more predictable effort) and defines projects as something with a fixed cost and time. Therefore, if a change occurs, other less important elements must be discarded to maintain the fixed amount. Software development is a relatively new field in comparison to the construction industry and was able to quickly adopt this new model. In contrast, effort estimation in the construction industry faces a long tradition of fee structures that is difficult to reinvent. This leaves a gap in research of algorithimic models that predict effort (and not construction cost) for consulting companies in the construction industry. From findings in previous literature, the models should reveal their predictive structure to decision makers and be based on internal company data. Review of the literature as a whole reveals a lack in analysis of overall profit improvement a model could provide to a business. This could pesuade industry adoption and provide a framework around how models should affect decisions.

@@

AS stated, the aim of this project is to use statistical and machine learning techniques to model the profitability of projects. Therefore, in order to optimise the model, several statistical and machine learning techniques will be tested and it is prudent to review the range of algorithms that have been successfully applied in the literature. These vary from simple methods such as Linear Regression to complex, deep learning Neural Networks. The previous section highlighted the use of Linear Regression, Neural Networks, SVMs and in one case Boosted Trees for prediction, however research on other business problems utilised a wider range of methods. These include Naiive Bayes, Random Forests, and machine learned Bayesian Networks. The following section presents each method's advantages and disadvatanges and an example of a past successful application.

@@

Linear regression is one of the simplest and most popular statistical prediction methods [@Hastie2009]. An equation for a line is learned, which is defined by a constant and a coefficient times the value of each variable. The coefficients are chosen to minimise the residual sum of squares (least squared) [@Hastie2009]. The reason this method is popular is that it is simple, provides stable predictions, and can be adapted for categorical variables [@Seng2010]. The integration of categorical variables is called Anaylis of Variance (ANOVA) Linear Regression, and the algorithm compares group mean variances within the categorical variables.

There are several drawbacks to Linear Regression. For example assumptions must be made about the structure of the data. Predictor variables are assumed to be linearly related to the response variable and the variables to be normally distributed. Many real world phenomena do not correspond to these assumptions which can make it difficult, sometimes impossible, to produce useful results [@Hastie2009][@Putler2012]. Furthermore, response variables must be continous numeric variables and if two or more predictor variables are correlated, the values of their coefficients are highly unstable. On the other hand, correlated variables do not impact the predictive accuracy of the model. A final drawback is that the variable coefficients cannot be used to compare variable importance unless the variables have all been standardised to one another (i.e. scaled to the same mean and standard deviation) [@Putler2012]. This limits the amount of insight that can be drawn from a Linear Regression model.

@@

![Plot of log odds vs. an 'x' variable and probability vs. the 'x' variable. To convert log odds to probability, an exponential transformation is first made followed by the transformation from odds to probabilitiy [@Lowry2016].](images/lit_review/logreg.JPG)

@@

The Naive Bayes method works by making conditional independence assumptions about the explanatory variables in order to greatly simplify probability calculations for the response variable (the response variable must be categorical). More specifically, it is assumed that each variable is conditionally independent given the respose class, without considering other predictor variables. Mathematically this means that each variable's probability contribution can be simply calculated by looking at the probability of a response class given a single predictor value. A probability of each response class can be calculated for each explanatory variable. These are then combined for each possible response class using Bayes theorem, and the response class with the highest probability is chosen [@Provost2013].

The advantages of this method are that the conditional independence assumption enables very fast calculations and predictions. The method can perform very well for real world tasks because the assumption of independence does not damage predictions significantly. This is because if multiple variables are related, the variables independently still direct the prediction in the correct direction. The correlated variables will double or triple their emphasis on the predicted response variable, however this does not necessarily impact classification accuracy as the final step is to simply choose the class with the *highest* probability [@Provost2013]. This is fine for ranking. On the other hand, the output probabilities are not realisitic, and the values themselves should not be used [@Caruana2006]. Another disadvantage is that in order to calculate the class probability from a numeric variable, a distribution must be assumed (often gaussian) which the numeric data may not fit into neatly [@Caruana2006]. Finally, Naive Bayes classifers are known to not perform well in binary classification [@Caruana2006].

@@

To combat the problems of low predictive accuracy and instability, ensemble tree methods were pioneered in the 1990's with success [@Breiman1996]. Three examples of these ensemble decision tree methods are boostrap aggregatiing (bagging), boosting, and Random Forests. Bagging creates multiple trees by sampling a different training data set (*with* replacement) for each tree and then combining tree results [@Breiman1996]. Random forests is an advanced form of bagging and builds on the idea of training multiple trees from the same data by sampling bootstrapped training sets with replacement. However, when creating each tree, a random subset of attributes (variables) are considered at each split. The reduced subset of attributes is resampled for each split in the tree. This allows dominant variables to be supressed for a fraction of the splits, allowing the algorithm to explore signals in weaker variables. It also prevents the trees from becoming too correlated [@Breiman2001a]. The boosted decision tree approach applies gradient descent theory to a series of decision trees. The trees are limited to a certain depth to maintain simplicity, and each tree models the residuals (or errors) of the preceding tree. The limited depth of each tree prevents overfitting at each stage and the combined result of up to thousands of trees is very powerful [@Elith2008].

@@

A Bayesian Network is a graphical probabilistic model that illustrates the conditional dependencies between variables in a data set. The model is visually represented by a DAG (directed acyclic graph) and is capable of linking the conditional dependency between any variable to another variable or via other variables. The links are graphically represented by arcs or edges [@Heckerman1998]. This kind of conditional relationship is bayesian, where the probabilities in one node are conditional upon values in nodes directed to it as well as preceding nodes. Bayesian networks have found success in combining deterministic models with observed data as well as expert knowledge. The statistical relationships between variables can be determined from separate sources and manually combined into a single model [@Kragt2009].

Bayesian networks are excellent at representing knowledge from different origins, and then visually communicating results well to decision makers [@Kragt2009]. Therefore Bayesian Networks have become an excellent decision support tool. Network relationships can be machine learned, however this is not widely included as part of the suite of machine learning methods and is usually recommeneded for modelling complex networks. Drawbacks of Bayesian Networks include their inability to handle continuous numeric variables (all continuous variables must be discretised). This ensures all variable relationships are categorical [@Kragt2009]. Popular applications of Bayesian Networks are modelling uncertainty in natural resource management and modelling complex business network structures such as airports [@Wu2013].

@@

This concept can be adapted to non linear boundaries by performing kernel transformations on prior variables as well as regression problems by introducing loss functions [@Auria2008]. Advantages of SVM's include the ability to model non linear relationships in data, and that no assumptions about the distributions of the data must be made as the model is non-probabilistic. They are also known to be robust over different samples and perform well in high-dimensional feature space [@Auria2008]. On the other hand, the output is in the form of distance to the boundary (as opposed to probability) and results are not transparent (similar to Neural Networks) [@Caruana2006]. Also, SVM's are known to be very slow to train and therefore often not suitable for industry purposes [@Auria2008]. Because SVM's perform well in high-dimensional space and can form non-linear rules they have been applied to protein classification (medicine) and text and image recognition [@Byun2002].

@@

The aim of this research project is to create a predictive model for the profitability of a company's consulting projects using their internal CRM data. The literature pertaining closely to the cost prediction topic has been reviewed in the first section of this chapter, however it is relevant to broaden the critique of the literature on advanced machine learning techniques that have been applied to general business problems. This is to research whether the broader literature found success in methods that cost estimation had not put to the test. The simpler statistical methods have been covered already, so the focus of this section are the machine learning prediction methods developed for business decisions. Popular applied topics include predicting stock fluctuation, customer churn analysis, fraud prediction, customer classification, market segment analysis, product success prediction, and recommendation systems (Seng & Chen, 2010). This section will first discuss the use and merits of the advanced machine learning methods applied to these problems, followed by a case study of employee-churn prediction.

@@

* profit

@@

* return per dollar as the measure of profitability

@@

results.core = readRDS("report_data/reg_lm_core.rds")
# n.samples(data = results.core$diff)

>>>>

opts_chunk$set(fig.width=6.7, dpi = 300, warning = FALSE, message = FALSE, echo = FALSE)
```

@@

# Table of Contents

# Table of Figures

@@

**ANOVA** - Analysis of Variance. A statistical test that can be applied to find whether there is a significant difference between the means of more than two groups.

@@

**ROC** - receiver Operating Characteristic. An ROC curve plots the performance of a binary classification algorithm its scoring threshold is varied.

@@

@Lovallo2003 call the practise of analysing a project based on project details and its unique complexities the 'inside view'. This contrasts with an 'outside view' where a subject is analysed by comparison to other subjects and disregarding specific subject details. For example, research has shown that if people make predictions about their skills after being exposed to an 'outside view' (a summary of other people's skills), their predictions are significantly more accurate. According to @Lovallo2003, this 'outside view' can be applied to complex projects, and stipulates that the details of the current project are ignored in favour of analysing outcomes of several similar projects. Mobilising the outside view in this way is called reference class forecasting [@Lovallo2003]. One method @Lovallo2003 recommend is to obtain correlation statistics from past similar projects - the correlation between the forecast cost and the actual cost. The correlation for the current project can then be estimated via a statistical model, which is then used to adjust the forecasted cost made by detailed analysis (the inside view) [@Flyvbjerg2011]. Managers rarely analyse data from similar projects even though it could significantly improve cost predictions [@Flyvbjerg2011]. This tendency led to a formalised introduction of 'outside view' methods into some institutions.

@@

In contrast to construction, IT projects are a relatively new practise where alternative contractural arangements have been trialed. Through experimentation with contracts, the Agile movement was developed. In summary, the method treats both cost and time as fixed quantities for a project, and any change or additional work can only be accommodated if another, less important requirement is excluded [@Badenfelt2011]. Furthermore, the project outcomes are continually adjusted ad revised based on frequent progress review points that assesses unexpected problems. This agile attitude has become ingrained in many IT practices and the flexible upfront arrangement with clients often forges better long term relationships [Cockburn2001].

@@

A statistical or machine learning model predicting a measure of project financial success was developed using the available project information and by trialing numerous machine learning algorithms and statistical methods. The company uses well established CRM software which readily provides simple output statistics, however, analytical capabilities were limited to:

@@

Fixed price cost estimation for complex projects is a difficult task and when performed incorrectly, can have devastating affects on businesses and governments. Many theories have delved into human nature and our tendency to optimistically assess our capabilities with respect to a given task. In this case, the optimism often results in negative business consequences affecting employee morale and the capacity of a company to flourish and grow. Over the past few decades, significant research has been dedicated to creating predictive models that present the 'outside view' of a project, by statistically comparing a new project to a collection of similar projects and their characteristics. It has been shown that these models improve the cost estimation, however industry uptake has not been successful. It has proven to be a challenging task persuading managers to use reference class forecasting models in practice and remove themselves from summaries based on intricate details of the completed project. This may be due to a number of reasons including lack of understanding, time, and lack of collaboration with industry in development of these models. This research examines the potential for a company's internally generated CRM data to be used in creating a reference class forecasting model. Familiarity with the data has the potential to influence managers' decisions in the company more, however its limitations must be understood in order to create a valid model.

@@

In addition to addressing cost estimation with a different kind of dataset and different industry sector, ensemble tree methods will be applied which have been minimally tested on this problem. To date, Linear Regression, Neural Networks, case based reasoning, and SVM's have been used to predict cost/effort even though ensemble tree methods can perform as well as Neural Networks in some cases and generally outperform Linear Regression. Another benefit of ensemble trees is the output types available, such as partial dependency plots and variable importance plots provide insight into the model's calculations. This is in contrast to Neural Networks and support vector machines (SVM's) which are 'black-box' predictors. Providing insight to a user is a major benefit in this case as past research has demonstrated issues with user trust and uptake of models, even if the models were successful. It is anticipated that using internal data only also improves user trust and creates tangible meaning for users. Blending multiple machine learning and statistical models was trialed which has not yet been applied to the cost estimation problem.

@@

@Elfaki2014 discusses how much detailed cost estimates for construction projects can vary from estimator (engineer) to estimator, which contributes to the lack of accuracy in predicting final cost of a project. He argues that an engineer's expertise is not documented or measured in any way and therefore their expertise and, in turn, their estimate is prone to subjectivity. @Shane2009 theorises that final cost can be subject to so many different unpredictable parameters, such as weather, unexpected ground conditions, duration or sub-contractor issues, that it is almost impossible to achieve an accurate cost prediction manually.

@@

@Elfaki2014 completed a study on cost estimation techniques and research from 2004 to 2014. They found that artificial Neural Networks and SVM's were the most common machine learning techniques used. He highlighted that these two machine learning techniques deal well with uncertainty, however lack technical justification to the decision maker (known as a black box). Neural networks are also time consuming to train, and must be re-trained and re-tested with each additional piece of data [@Kim2004]. Nevertheless, Neural Networks and SVM's received a significant amount of attention in the 1990's for their ability to accurately predict construction costs with limited detailed information [@Shin2015][@Kim2004].

@@

Despite Neural Networks typically outperforming Linear Regression models, it is interesting to note some of the outcomes of regression studies in cost estimation. Different studies tended to produce a wide range of variables that contribute at fluctuating levels in the models. This is likely because construction projects are multidisciplinary and involve a range of different parties such as the clients, consultants, contractors and suppliers, each of which contribute varying levels of input [@Chan2005]. @Chan2005's study on 87 building projects in Singapore included special, complex projects. Therefore, variables such as contractor's specialised skills, who the client was (public vs. private), the client level of experience, and the contractor's financial management highly influenced final cost of the project. Other studies found variables such as project complexity, duration, team experience, information availablility, site requirements, and labour climate to contribute most [@Akintoye2000][@Trost2003][@Pinto1988]. As these studies obtained information from questionnaires, the available variables also largely depended on the questions in the questionnaires, and therefore each study produced a range of different important variables. Despite the differences in qualitative outcomes, the studies all generally reported their mean absolute error or a similar metric from the tested models. Similar result structures are useful, however there is a gap in the literature in how to translate the improved errors from the model into a well justified argument for industry to reduce losses by integrating a parametric model into their business.

@@

1. Storing a collection of projects with key variable values
2. Once a new project or case arrives, similar cases are retrieved. This can be achieved either algorithmically with a distance function applied to the variables, or manually by users reviewing past cases. The nearest neighbour algorithm has been used to find the distance between cases in previous studies. It is programmed to calculate the Euclidean distance in n-dimensional space between cases, where each variable is a dimension [@Kumar2007].
3. The cost of the new project is estimated by extrapolating characteristics of the similar cases to the new case. This can also be done either algorithmically or manually by an experienced decision maker.

@@

In summary, research into cost estimation in the construction industry has trialed a variety of methods ranging from detailed manual analysis, algorithmic models, and case based reasoning. Neural networks generally outperformed all other systematic methods, such as regression and case based reasoning, but had the significant disadvantage of being a black box. Alternative methods such as Boosted Trees have the potential to perform as well as Neural Networks and provide insight into the structure of the algorithm. This is true for CBR as well, although it has not been shown to predict as accurately as Neural Networks. There exists a gap in the research in ways to intelligently combine CBR with machine learning methods that could predict well, provide insight to the structure, and engage the user in reviewing similar projects. Also, as stated previously, there exists a gap in the application of cost estimating in the construction industry to smaller components of the project, such as consulting and contracting companies who each face a similar problem in estimating their fixed fee. These companies could mobilise information in their internal project databases, as opposed to the current body of literature that used data from dozens of companies.

@@

Effort estimation is indeed a problem, and was demonstrated by a review of 6 surveys of IT projects between 1984 and 1994 by @Molokken2003. It revealed that 60-80% of IT projects encountered effort and/or schedule overruns, where the average overrun was by 30 to 40%. A survey of software managers asked whether they saw effort estimation as a problem, and 91% responded 'yes', while only 9% answered 'no' [@Moores1992]. The most popular response for the supposed reason for cost overruns was over-optimistic estimates (51%), which falls squarely within the theory of optimism bias, or the 'inside view' [@Flyvbjerg2011]. Given this widespread problem in the industry, plenty of research has been dedicated to the problem of effort prediction, with methods that can be categorised similarly to the construction industry: detailed analysis, algorithmic modelling, and case based reasoning.

@@

Similar to the construction industry, expert judgment or detailed analysis, is the most widely practised method for effort estimation [@Shepperd1996][@Molokken2003]. This is despite many years of research being dedicated to developing algorithmic models and industry tools that, in the research context, outperform expert judgment. A survey by @Moores1992 found that software tools assisting a software project were in significant use, but only the project planning tools, not the cost estimating tools. This suggests a mental link for project managers to price jobs based on *details* of the software project as opposed to taking an outside view. Another study by @Heemstra1992 found that there was no evidence estimation accuracy improved when estimation tools were used. @Bergeron1992 similarly found jobs that used algorithmic models were actually associated with less accurate estimates. However, this trend may be coincidental due to a lack of cases where estimation tools were used [@Molokken2003]. The scarcity of real project evidence that algorithmic estimation tools improve estimations and an unintuitive mental jump could be reasons contributing to detailed expert analysis remaining the most widespread technique for effort estimation. This trend mirrors what was found in the construction industry.

@@

As alluded to previously, studies on effort estimation as opposed to project cost estimation are concentrated in the software industry. This leaves a gap in studies for effort estimation in other industries such as the construction industry. Infrastructure projects do not have a function point variable and would need to utilise other information such as expected total project cost (which is notoriously hard to predict based on the previous section), client characteristics, and other project characteristics.

@@

In the IT industry, CBR applications have been researched using similar methodologies to the construction industry. The procedure is described as algorithmically finding similar projects using measurements of Euclidean distance in n-dimensional space, where each dimension corresponds to a variable [@Shepperd1996]. Then, the similar cases are used to predict effort for the new case. In @Shepperd1996's study, Linear Regression was employed.

@@

Research into effort estimation in the software industry has followed a similar path to cost prediction studies on construction projects. Expert judgment via detailed analysis is still the predominant method for effort/cost estimation. Case based reasoning and algorithmic methods such as Neural Networks have found some success in effort estimation (at similar levels) but have not been successfully adopted by industry. It was noted that case based reasoning resonated more with decision makers because its predictions could be better justified by the model. The visibility of model structure is very important for future work if the results are to be translated to industry. Effort estimation remains a problem, however agile methods of project management and effort management are changing the contractual approach of consulting in the software industry [@Badenfelt2011]. This innovative method keeps delivery times short (and therefore more predictable effort) and defines projects as something with a fixed cost and time. Therefore, if a change occurs, other less important elements must be discarded to maintain the fixed amount. Software development is a relatively new field in comparison to the construction industry and was able to quickly adopt this new model. In contrast, effort estimation in the construction industry faces a long tradition of fee structures that is difficult to reinvent. This leaves a gap in research of algorithmic models that predict effort (and not construction cost) for consulting companies in the construction industry. From findings in previous literature, the models should reveal their predictive structure to decision makers and be based on internal company data. Review of the literature as a whole reveals a lack in analysis of overall profit improvement a model could provide to a business. This could persuade industry adoption and provide a framework around how models should affect decisions.

@@

AS stated, the aim of this project is to use statistical and machine learning techniques to model the profitability of projects. Therefore, in order to optimise the model, several statistical and machine learning techniques will be tested and it is prudent to review the range of algorithms that have been successfully applied in the literature. These vary from simple methods such as Linear Regression to complex, deep learning Neural Networks. The previous section highlighted the use of Linear Regression, Neural Networks, SVMs and in one case Boosted Trees for prediction, however research on other business problems utilised a wider range of methods. These include Naive Bayes, Random Forests, and machine learned Bayesian Networks. The following section presents each method's advantages and disadvantanges and an example of a past successful application.

@@

Linear regression is one of the simplest and most popular statistical prediction methods [@Hastie2009]. An equation for a line is learned, which is defined by a constant and a coefficient times the value of each variable. The coefficients are chosen to minimise the residual sum of squares (least squared) [@Hastie2009]. The reason this method is popular is that it is simple, provides stable predictions, and can be adapted for categorical variables [@Seng2010]. The integration of categorical variables is called Analysis of Variance (ANOVA) Linear Regression, and the algorithm compares group mean variances within the categorical variables.

There are several drawbacks to Linear Regression. For example assumptions must be made about the structure of the data. Predictor variables are assumed to be linearly related to the response variable and the variables to be normally distributed. Many real world phenomena do not correspond to these assumptions which can make it difficult, sometimes impossible, to produce useful results [@Hastie2009][@Putler2012]. Furthermore, response variables must be continuous numeric variables and if two or more predictor variables are correlated, the values of their coefficients are highly unstable. On the other hand, correlated variables do not impact the predictive accuracy of the model. A final drawback is that the variable coefficients cannot be used to compare variable importance unless the variables have all been standardised to one another (i.e. scaled to the same mean and standard deviation) [@Putler2012]. This limits the amount of insight that can be drawn from a Linear Regression model.

@@

![Plot of log odds vs. an 'x' variable and probability vs. the 'x' variable. To convert log odds to probability, an exponential transformation is first made followed by the transformation from odds to probability [@Lowry2016].](images/lit_review/logreg.JPG)

@@

The Naive Bayes method works by making conditional independence assumptions about the explanatory variables in order to greatly simplify probability calculations for the response variable (the response variable must be categorical). More specifically, it is assumed that each variable is conditionally independent given the response class, without considering other predictor variables. Mathematically this means that each variable's probability contribution can be simply calculated by looking at the probability of a response class given a single predictor value. A probability of each response class can be calculated for each explanatory variable. These are then combined for each possible response class using Bayes theorem, and the response class with the highest probability is chosen [@Provost2013].

The advantages of this method are that the conditional independence assumption enables very fast calculations and predictions. The method can perform very well for real world tasks because the assumption of independence does not damage predictions significantly. This is because if multiple variables are related, the variables independently still direct the prediction in the correct direction. The correlated variables will double or triple their emphasis on the predicted response variable, however this does not necessarily impact classification accuracy as the final step is to simply choose the class with the *highest* probability [@Provost2013]. This is fine for ranking. On the other hand, the output probabilities are not realistic, and the values themselves should not be used [@Caruana2006]. Another disadvantage is that in order to calculate the class probability from a numeric variable, a distribution must be assumed (often Gaussian) which the numeric data may not fit into neatly [@Caruana2006]. Finally, Naive Bayes classifiers are known to not perform well in binary classification [@Caruana2006].

@@

To combat the problems of low predictive accuracy and instability, ensemble tree methods were pioneered in the 1990's with success [@Breiman1996]. Three examples of these ensemble decision tree methods are boostrap aggregating (bagging), boosting, and Random Forests. Bagging creates multiple trees by sampling a different training data set (*with* replacement) for each tree and then combining tree results [@Breiman1996]. Random forests is an advanced form of bagging and builds on the idea of training multiple trees from the same data by sampling bootstrapped training sets with replacement. However, when creating each tree, a random subset of attributes (variables) is considered at each split. The reduced subset of attributes is resampled for each split in the tree. This allows dominant variables to be suppressed for a fraction of the splits, allowing the algorithm to explore signals in weaker variables. It also prevents the trees from becoming too correlated [@Breiman2001a]. The boosted decision tree approach applies gradient descent theory to a series of decision trees. The trees are limited to a certain depth to maintain simplicity, and each tree models the residuals (or errors) of the preceding tree. The limited depth of each tree prevents overfitting at each stage and the combined result of up to thousands of trees is very powerful [@Elith2008].

@@

A Bayesian Network is a graphical probabilistic model that illustrates the conditional dependencies between variables in a data set. The model is visually represented by a DAG (directed acyclic graph) and is capable of linking the conditional dependency between any variable to another variable or via other variables. The links are graphically represented by arcs or edges [@Heckerman1998]. This kind of conditional relationship is Bayesian, where the probabilities in one node are conditional upon values in nodes directed to it as well as preceding nodes. Bayesian networks have found success in combining deterministic models with observed data as well as expert knowledge. The statistical relationships between variables can be determined from separate sources and manually combined into a single model [@Kragt2009].

Bayesian networks are excellent at representing knowledge from different origins, and then visually communicating results well to decision makers [@Kragt2009]. Therefore Bayesian Networks have become an excellent decision support tool. Network relationships can be machine learned, however this is not widely included as part of the suite of machine learning methods and is usually recommended for modelling complex networks. Drawbacks of Bayesian Networks include their inability to handle continuous numeric variables (all continuous variables must be discretised). This ensures all variable relationships are categorical [@Kragt2009]. Popular applications of Bayesian Networks are modelling uncertainty in natural resource management and modelling complex business network structures such as airports [@Wu2013].

@@

This concept can be adapted to nonlinear boundaries by performing kernel transformations on prior variables as well as regression problems by introducing loss functions [@Auria2008]. Advantages of SVM's include the ability to model nonlinear relationships in data, and that no assumptions about the distributions of the data must be made as the model is non-probabilistic. They are also known to be robust over different samples and perform well in high-dimensional feature space [@Auria2008]. On the other hand, the output is in the form of distance to the boundary (as opposed to probability) and results are not transparent (similar to Neural Networks) [@Caruana2006]. Also, SVM's are known to be very slow to train and therefore often not suitable for industry purposes [@Auria2008]. Because SVM's perform well in high-dimensional space and can form non-linear rules they have been applied to protein classification (medicine) and text and image recognition [@Byun2002].

@@

The aim of this research project is to create a predictive model for the profitability of a company's consulting projects using their internal CRM data. The literature pertaining closely to the cost prediction topic has been reviewed in the first section of this chapter, however it is relevant to broaden the critique of the literature on advanced machine learning techniques that have been applied to general business problems. This is to research whether the broader literature found success in methods that cost estimation had not put to the test. The simpler statistical methods have been covered already, so the focus of this section is the machine learning prediction methods developed for business decisions. Popular applied topics include predicting stock fluctuation, customer churn analysis, fraud prediction, customer classification, market segment analysis, product success prediction, and recommendation systems (Seng & Chen, 2010). This section will first discuss the use and merits of the advanced machine learning methods applied to these problems, followed by a case study of employee-churn prediction.

@@

* Profit

@@

* Return per dollar as the measure of profitability

@@

results.core = readRDS("reg_lm_core.rds")
# n.samples(data = results.core$diff)
